Stochastic Variance-Reduced Policy Gradient

### 2018-02-09 svrpg.tex
Fix of typos

The main limitation of plain policy gradient is the high variance of these estimators.
The na\"ive approach of increasing the batch size is not an option in RL due to the high cost of collecting samples, \ie by interacting with the environment.
For this reason, literature has focused on the introduction of baselines (\ie functions $b : \mathcal{S} \times \mathcal{A} \to \realspace$) aiming to reduce the variance~\citep[\eg][]{williams1992simple,Peters2008reinf,Thomas2017actionbaseline,wu2018variance}, see Appendix~\ref{A:gradient_estimators} for a formal definition of $b$.
These baselines are usually designed to minimize the variance of the gradient estimate, but even them need to be estimated from data, partially reducing their effectiveness.
On the other hand, there has been a surge of recent interest in variance reduction techniques for gradient optimization in supervised learning.
Although these techniques have been mainly derived for finite-sum problems, we will show in Section~\ref{sec:alg} how they can be used in RL.
In particular, we will show that the proposed SVRPG algorithm can take the best of both worlds (\ie SL and RL) since it can be plugged onto a policy gradient estimate using baselines.
The next section has the aim to describe variance reduction techniques for finite-sum problems. In particular, we will present the SVRG algorithm that is at the core of this work.

@@

The idea of SVRG (Algorithm~\ref{alg:svrg}) is to alternate full and stochastic gradient updates. 
Each $m = \mathcal{O}(N)$ iterations, a snapshot $\widetilde{\vtheta}$ of the current parameter is saved together with its full gradient $\nabla f(\widetilde{\vtheta}) = \frac{1}{N} \sum_i \nabla z(x_i|\widetilde{\vtheta})$.
Between snapshots, the parameter is updated with $\blacktriangledown f(\vtheta)$, a gradient estimate corrected using stochastic gradient. For any $t \in \{0,\ldots,m-1\}$:
\begin{equation}\label{E:svrg.gradient.correction}
        \blacktriangledown f(\vtheta_{t}) := v_t = \nabla f(\wt{\vtheta}) + \nabla z(x | \vtheta_t) - \nabla z(x | \wt{\vtheta}),
\end{equation} 
where $x$ is sampled uniformly at random from $\mathcal{D}_N$ (\ie $x \sim \mathcal{U}(\mathcal{D}_N)$).
Note that $t=0$ corresponds to a FG step (\ie $\blacktriangledown f(\vtheta_0) = \nabla f(\wt{\vtheta})$) since $\vtheta_0 := \wt{\vtheta}$.
The corrected gradient $\blacktriangledown f(\vtheta)$ is an unbiased estimate of $\nabla f(\vtheta)$, and it is able to control the variance introduced by sampling even with a fixed step size, achieving a linear convergence rate without resorting to plain full gradient.

@@

\vspace{-0.05in}
\section{SVRG in Reinforcement Learning}\label{sec:alg}
\vspace{-0.05in}
In online RL problems, the usual approach is to tune the batch size of SG to find the optimal trade-off between variance and speed.
Recall that, compared to SL, the samples are not fixed in advance but we need to collect them at each policy change.
Since this operation may be costly, we would like to minimize the number of interactions with the environment.
For these reasons, we would like to apply SVRG to RL problems in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence.
However, a direct application of SVRG to RL is not possible due to the following issues:
% In on-line policy gradient problems, SG is not really a choice, but is dictated by the necessity of interacting with an unknown environment.\todopir{I do not understand the meaning of this sentence. I would say that SG is not a choice by referring to papers that uses batches}
% We would like to apply SVRG to the policy gradient framework in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence. In this scenario, the function $f(\vtheta)$ to optimize is $J(\vtheta)$ and the (implicit) dataset $\mathcal{D}$ is the set of all possible trajectories $\mathcal{T}$. Compared to the typical finite sum optimization scenario, we have the following additional challenges:
\begin{description}
        \item[Non-concavity:] the objective function $J(\vtheta)$ is typically non-concave.
        \item[Infinite dataset:] the RL optimization cannot be expressed as a finite-sum problem. The objective function is an expected value over the trajectory density $p_{\vtheta}(\tau)$ of the total discounted reward, for which we would need an infinite dataset.
%    \item[Approximation:] the dataset $\mathcal{T}$ can be infinite, but we can only sample a finite number of trajectories;
        \item[Non-stationarity:] the distribution of the samples changes over time. In particular, the value of the policy parameter $\vtheta$ influences the sampling process.
\end{description}
To deal with non-concavity, we require $J(\vtheta)$ to be $L$-smooth, which is a reasonable assumption for common policy classes such as Gaussian\footnote{See Appendix~\ref{app:gauss} for more details on the Gaussian policy case.} and softmax~\citep[\eg][]{Furmston2012unifying,pirotta2015lipschitz}.
Because of the infinite dataset, we can only rely on an estimate of the full gradient.
\citet{harikandeh2015stopwasting} analysed this scenario under the assumptions of $z$ being concave, showing that SVRG is robust to an inexact computation of the full gradient. In particular, it is still possible to recover the original convergence rate if the error decreases at an appropriate rate. In Section \ref{sec:conv} we will show how the estimation accuracy impacts on the convergence results with a non-concave objective.
Finally, the non-stationarity of the optimization problem introduces a bias into the SVRG estimator in Eq.~\eqref{E:svrg.gradient.correction}.
To overcome this limitation we employ importance weighting~\citep[\eg][]{rubinstein1981simulation,precup2000eligibility} to correct the distribution shift.
%Finally, we employ importance weighting~\citep[\eg]{rubinstein1981simulation,precup2000eligibility} to guarantee proper sampling of trajectories.

@@

\textbf{On-policy setting.}
Consider a policy $\pol$ and let $\mathcal{D}_N = \{\tau_i\}_{i=1}^N$ be a dataset collected using policy $\pol$.
The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
        \left(\sum_{h=0}^{H}\nabla \log \pol(z_h^n) \right)\left(\sum_{h=0}^{H}\gamma^h 
%        \Reward(z_h^n)
        r_h^n
- b(z_h^n)\right)
}_{g(\tau_n|\vtheta):=\nabla \log p(\tau_n|\vtheta)\Reward(\tau_n)}
,
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h^n$ is the reward actually collected at time $h$ from trajectory $\tau^n$ and $b : \mathcal{S}\times\mathcal{A} \to \realspace$~\citep[\eg][]{Thomas2017actionbaseline}.
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance \cite{zhao2011analysis} while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
        \sum_{h=0}^{H}\left(\sum_{k=0}^{h} \nabla \log \pol (z_h^n) \right)\left(\gamma^h 
%        \Reward(z_h^n)
        r_h^n
        - b(z^n_h)\right)
}_{g(\tau_n|\vtheta)}.
\end{align*}
G(PO)MDP can be seen as a more efficient implementation of the REINFORCE algorithm. 
In fact the latter does not perform an optimal credit assignment, since it ignores that the reward at time $t$ does not depend on the action performed after time $t$.
G(PO)MDP overcomes this issue taking into account the causality of rewards in the REINFORCE definition of policy gradient.

@@

Next we have:

@@

where separate histories are kept for estimated first moments $\kappa_{FG},\kappa_{IS}$ and estimated second moments $\nu_{FG},\nu_{IS}$.
The meta-parameters $\alpha,\beta_1,\beta_2$ are constant and set to default values or with minor manual tuning (see table \ref{table:metaparams}). Note that we double $\alpha$ for the snapshot ADAM, since we can rely on a larger number of trajectories ($N$ instead of $B$) to control the variance.

@@

\section{Experimental Details}\label{app:exp}
We describe the RL tasks of Section \ref{sec:exp} in more detail:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : an inverted pendulum mounted on a cart must be kept standing by moving the cart backward or forward ;4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminates when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
    \item \emph{Mujoco Swimmer}: a snake-like robot immersed in a fluid must move forward; 13-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2]{a}^2$. The episodes terminates when the number of time steps T is greater than 500.
    \item \emph{Mujoco Half Cheetah}: a planar biped robot must move forward; 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2]{a}^2$. The episodes terminates when the number of time steps T is greater than 500.
\end{enumerate}

>>>>

The main limitation of plain policy gradient is the high variance of these estimators.
The na\"ive approach of increasing the batch size is not an option in RL due to the high cost of collecting samples, \ie by interacting with the environment.
For this reason, literature has focused on the introduction of baselines (\ie functions $b : \mathcal{S} \times \mathcal{A} \to \realspace$) aiming to reduce the variance~\citep[\eg][]{williams1992simple,Peters2008reinf,Thomas2017actionbaseline,wu2018variance}, see Appendix~\ref{A:gradient_estimators} for a formal definition of $b$.
These baselines are usually designed to minimize the variance of the gradient estimate, but even them need to be estimated from data, partially reducing their effectiveness.
On the other hand, there has been a surge of recent interest in variance reduction techniques for gradient optimization in supervised learning.
Although these techniques have been mainly derived for finite-sum problems, we will show in Section~\ref{sec:alg} how they can be used in RL.
In particular, we will show that the proposed SVRPG algorithm can take the best of both worlds (\ie SL and RL) since it can be plugged into a policy gradient estimate using baselines.
The next section has the aim to describe variance reduction techniques for finite-sum problems. In particular, we will present the SVRG algorithm that is at the core of this work.

@@

The idea of SVRG (Algorithm~\ref{alg:svrg}) is to alternate full and stochastic gradient updates. 
Each $m = \mathcal{O}(N)$ iterations, a snapshot $\widetilde{\vtheta}$ of the current parameter is saved together with its full gradient $\nabla f(\widetilde{\vtheta}) = \frac{1}{N} \sum_i \nabla z(x_i|\widetilde{\vtheta})$.
Between snapshots, the parameter is updated with $\blacktriangledown f(\vtheta)$, a gradient estimate corrected using stochastic gradient. For any $t \in \{0,\ldots,m-1\}$:
\begin{equation}\label{E:svrg.gradient.correction}
        \blacktriangledown f(\vtheta_{t}) := v_t = \nabla f(\wt{\vtheta}) + \nabla z(x | \vtheta_t) - \nabla z(x | \wt{\vtheta}),
\end{equation} 
where $x$ is sampled uniformly at random from $\mathcal{D}_N$ (\ie $x \sim \mathcal{U}(\mathcal{D}_N)$).
Note that $t=0$ corresponds to a FG step (\ie $\blacktriangledown f(\vtheta_0) = \nabla f(\wt{\vtheta})$) since $\vtheta_0 := \wt{\vtheta}$.
The corrected gradient $\blacktriangledown f(\vtheta)$ is an unbiased estimate of $\nabla f(\vtheta)$, and it is able to control the variance introduced by sampling even with a fixed step size, achieving a linear convergence rate without resorting to a plain full gradient.

@@

\vspace{-0.05in}
\section{SVRG in Reinforcement Learning}\label{sec:alg}
\vspace{-0.05in}
In online RL problems, the usual approach is to tune the batch size of SG to find the optimal trade-off between variance and speed.
Recall that, compared to SL, the samples are not fixed in advance but we need to collect them at each policy change.
Since this operation may be costly, we would like to minimize the number of interactions with the environment.
For these reasons, we would like to apply SVRG to RL problems in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence.
However, a direct application of SVRG to RL is not possible due to the following issues:
% In on-line policy gradient problems, SG is not really a choice, but is dictated by the necessity of interacting with an unknown environment.\todopir{I do not understand the meaning of this sentence. I would say that SG is not a choice by referring to papers that use batches}
% We would like to apply SVRG to the policy gradient framework in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence. In this scenario, the function $f(\vtheta)$ to optimize is $J(\vtheta)$ and the (implicit) dataset $\mathcal{D}$ is the set of all possible trajectories $\mathcal{T}$. Compared to the typical finite sum optimization scenario, we have the following additional challenges:
\begin{description}
        \item[Non-concavity:] the objective function $J(\vtheta)$ is typically non-concave.
        \item[Infinite dataset:] the RL optimization cannot be expressed as a finite-sum problem. The objective function is an expected value over the trajectory density $p_{\vtheta}(\tau)$ of the total discounted reward, for which we would need an infinite dataset.
%    \item[Approximation:] the dataset $\mathcal{T}$ can be infinite, but we can only sample a finite number of trajectories;
        \item[Non-stationarity:] the distribution of the samples changes over time. In particular, the value of the policy parameter $\vtheta$ influences the sampling process.
\end{description}
To deal with non-concavity, we require $J(\vtheta)$ to be $L$-smooth, which is a reasonable assumption for common policy classes such as Gaussian\footnote{See Appendix~\ref{app:gauss} for more details on the Gaussian policy case.} and softmax~\citep[\eg][]{Furmston2012unifying,pirotta2015lipschitz}.
Because of the infinite dataset, we can only rely on an estimate of the full gradient.
\citet{harikandeh2015stopwasting} analysed this scenario under the assumptions of $z$ being concave, showing that SVRG is robust to an inexact computation of the full gradient. In particular, it is still possible to recover the original convergence rate if the error decreases at an appropriate rate. In Section \ref{sec:conv} we will show how the estimation accuracy impacts on the convergence results with a non-concave objective.
Finally, the non-stationarity of the optimization problem introduces a bias into the SVRG estimator in Eq.~\eqref{E:svrg.gradient.correction}.
To overcome this limitation we employ importance weighting~\citep[\eg][]{rubinstein1981simulation,precup2000eligibility} to correct the distribution shift.
%Finally, we employ importance weighting~\citep[\eg]{rubinstein1981simulation,precup2000eligibility} to guarantee proper sampling of trajectories.

@@

\textbf{On-policy setting.}
Consider a policy $\pol$ and let $\mathcal{D}_N = \{\tau_i\}_{i=1}^N$ be a dataset collected using policy $\pol$.
The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
        \left(\sum_{h=0}^{H}\nabla \log \pol(z_h^n) \right)\left(\sum_{h=0}^{H}\gamma^h 
%        \Reward(z_h^n)
        r_h^n
- b(z_h^n)\right)
}_{g(\tau_n|\vtheta):=\nabla \log p(\tau_n|\vtheta)\Reward(\tau_n)}
,
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h^n$ is the reward actually collected at time $h$ from trajectory $\tau^n$ and $b : \mathcal{S}\times\mathcal{A} \to \realspace$~\citep[\eg][]{Thomas2017actionbaseline}.
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance \cite{zhao2011analysis} while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
        \sum_{h=0}^{H}\left(\sum_{k=0}^{h} \nabla \log \pol (z_h^n) \right)\left(\gamma^h 
%        \Reward(z_h^n)
        r_h^n
        - b(z^n_h)\right)
}_{g(\tau_n|\vtheta)}.
\end{align*}
G(PO)MDP can be seen as a more efficient implementation of the REINFORCE algorithm. 
In fact, the latter does not perform an optimal credit assignment since it ignores that the reward at time $t$ does not depend on the action performed after time $t$.
G(PO)MDP overcomes this issue taking into account the causality of rewards in the REINFORCE definition of policy gradient.

@@

Next, we have:

@@

where separate histories are kept for estimated first moments $\kappa_{FG},\kappa_{IS}$ and estimated second moments $\nu_{FG},\nu_{IS}$.
The meta-parameters $\alpha,\beta_1,\beta_2$ are constant and set to default values or with minor manual tuning (see table \ref{table:metaparams}). Note that we double $\alpha$ for the snapshot ADAM since we can rely on a larger number of trajectories ($N$ instead of $B$) to control the variance.

@@

\section{Experimental Details}\label{app:exp}
We describe the RL tasks of Section \ref{sec:exp} in more detail:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : an inverted pendulum mounted on a cart must be kept standing by moving the cart backward or forward ;4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminate when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
    \item \emph{Mujoco Swimmer}: a snake-like robot immersed in a fluid must move forward; 13-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2]{a}^2$. The episodes terminate when the number of time steps T is greater than 500.
    \item \emph{Mujoco Half Cheetah}: a planar biped robot must move forward; 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2]{a}^2$. The episodes terminate when the number of time steps T is greater than 500.
\end{enumerate}
